Reproducible, flexible and high throughput data extraction from primary literature: The metaDigitise R package

### 2017-11-03 docs/manuscript.tex
Writings MS

\title{Reproducible, flexible and high throughput data extraction from primary literature for meta-analyses: The metaDigitise R package}
\author{Joel L. Pick, Shinichi Nakagawa \& Daniel W.A. Noble}
% \date{}

@@

Keywords: meta-analysis, data extraction, R, reproducibility, figures, images, summary statistics

@@

\section{Functionality}
There is one main function in the package metaDigitise, which interactively takes the user through the process of extracting figures from graphs.

\subsection{Bulk Digitisation}
metaDigitise was created with the idea that the user would have multiple images to extract from. It therefore operates in the same way whether the user has one or multiple images. It assumes that the user has put all the figures needed for extraction in one parent directory. It first creates a new directory `caldat' inside this parent directory, in which calibration files are saved (see below).

If the process of data extraction has been finished, or the user wants to reimport summary or processed data ...

\subsection{Figure Rotation}
Figure may have been extracted from old publications, for example from scanned images, and so are not perfectly orientated on the image. This will make the calibration of the points in the figure from the image problematic. metaDigitise allows users to rotate the image. By clicking two points on the x-axis, metaDigitse calculates the angle needed to rotate the image so the x-axis is horizontal, and rotates it. (Figure \ref{fig:rotate}A,B)

\subsection{Calibration}
metaDigitise requires the user to calibrate the axes in the figure. To do this the user is required to click on two known points on the axis in question, and then enter the value of those points in the figure (Figure \ref{fig:calibrate}). Using this information, metaDigitise then calculates the value of any clicked points in terms of the figure axes. In the case of mean and error plots and box plots, it calibrates only the y axis (assuming the x axis is redundant). For scatter plots and histograms both axes are calibrated.

@@

\subsection{Plot Types}
metaDigitise recognises 4 main types of plot; Mean and error plots, box plots, scatter plots and histograms, shown in Figure \ref{fig:plot_type}.

@@

\subsection{Summary Data}
From all plot types, metaDigitise summarises the data form the figure to a mean, standard deviation and sample size, for each identified group. These are the summary statistical needed to create all relevant statistics for a meta-analysis. In the case of scatter plots, metaDigitise also returns the correlation coefficient between the points within each identified group.

%%% Equations for mean and sd

%% pooled stats: 
% \begin{align}
% \mu _{X}&={\frac {\sum _{i}{N_{X_{i}}\mu _{X_{i}}}}{\sum _{i}{N_{X_{i}}}}}
% \\
% \sigma _{X}&={\sqrt {{\frac {1}{\sum _{i}{N_{X_{i}}-1}}}\left(\sum _{i}{\left[(N_{X_{i}}-1)\sigma _{X_{i}}^{2}+N_{X_{i}}\mu _{X_{i}}^{2}\right]}-\left[\sum _{i}{N_{X_{i}}}\right]\mu _{X}^{2}\right)}}
% \end{align}

\section{Directory Structure and Reproducibility}

The `metaDigitise` package is quite flexible. Users can extract single figures (if this is all they have) using the `metaDigitise` function with a path name to the directory with the file. However, often many figures need extracting from a single paper or set of papers. `metaDigitise` will also handle these situations seamlessly by simply cycling through all figures within a directory. This is useful because it expedites digitising figures as it prevents users from having to constantly specify the directories and / or paths where files are stored. `metaDigitise` essentially will bring up each figure within a folder automatically and allow the user to click and enter the relevant information about a figure as they go. This information is then all stored in a data frame or list at the end of the process, saving quite a bit of time. Users can stop mid-way through a folder by simply exiting after the last plot they have digitised.  The data from completed figures will automatically be written to the `caldat/` folder for later use and editing, should the user need to do this.

`metaDigitise` can work on a directory with figures (currently .png, .jpg, .tiff, .pdf images can be used) from many different papers and that are of different types. However, users can get creative in how they set up the directories of figures to facilitate extraction. For example, one might have 3–4 figures from a single paper that need extracting and the user may want to focus on a single paper at a time while the information about a paper is on hand. This could be done by simply setting up a file structure as follows and then using `metaDigitise` with path names (i.e., directory) for each papers folder:

Nonetheless, how users set up their directory is really up to them. However, it is important for users to think carefully about reproducibility at this stage. Would they like to share the entire project folder with colleagues? or would they prefer to simply share the image folder? The answers to these questions are important because relative path names are stored in `metaDigitise` objects, meaning that the directory structure (currently at least) needs to be the same for colleagues to re-load previously digitised objects. Therefore, the working directory needs to be set to either the "main project directory" or the "FiguresToExtract" directory and relative path names (e.g., dir = "./FiguresToExtract/" if setting the working directory to "Main project directory") used as opposed to absolute paths. This facilitates reproducibility should colleagues want to see the digitisations or integrate previously digitised figures with newly updated figures. All users need to provide them with is the project or image folder containing the caldat/ folder and the images.

\subsection{Plotting data extraction}

%check extraction

\subsection{Editing data extraction}

%Including addition of N later

\section{Example}
%% scatterplot from airquality data

>>>>

\title{Reproducible, flexible and high throughput data extraction from primary literature: The metaDigitise R package}
\author{Joel L. Pick, Shinichi Nakagawa \& Daniel W.A. Noble}
% \date{}

@@

Keywords: meta-analysis, comparative analysis, data extraction, R, reproducibility, figures, images, summary statistics

@@

\section*{Directory Structure and Reproducibility}

The `metaDigitise` package is quite flexible. There is one main function in the package, metaDigitise, which interactively takes the user through the process of extracting figures from graphs. `metaDigitise` can work on a directory with figures (currently .png, .jpg, .tiff, .pdf images can be used) from many different papers and that are of different types (see below). However, users can get creative in how they set up the directories of figures to facilitate extraction. For example, one might have 3–4 figures from a single paper that need extracting and the user may want to focus on a single paper at a time while the information about a paper is on hand. This could be done by simply setting up a file structure as follows and then using `metaDigitise` with path names (i.e., directory) for each papers folder:

Nonetheless, how users set up their directory is really up to them. However, it is important for users to think carefully about reproducibility at this stage. Would they like to share the entire project folder with colleagues? or would they prefer to simply share the image folder? The answers to these questions are important because relative path names are stored in `metaDigitise` objects, meaning that the directory structure (currently at least) needs to be the same for colleagues to re-load previously digitised objects. Therefore, the working directory needs to be set to either the "main project directory" or the "FiguresToExtract" directory and relative path names (e.g., users should specify dir = "./FiguresToExtract/" if setting the working directory to "Main project directory") used as opposed to absolute paths. This facilitates reproducibility should colleagues want to see the digitisations or integrate previously digitised figures with newly digitised figures. All users need to provide them with is the project or image folder.

\section*{High-throughput Digitisation of Figure(s) within a Directory}
metaDigitise was created with the idea that the user would likely have multiple images to extract from. It therefore operates in the same way whether the user has one or multiple images. It assumes that the user has put all the figures needed for extraction in one parent directory (see above). Users can extract single figures (if this is all they have) using the `metaDigitise` function with a path name to the directory with the file. However, often many figures need extracting from a single paper or set of papers. `metaDigitise` will also handle these situations seamlessly by simply cycling through all figures within a directory. This is useful because it expedites digitising figures by preventing users from having to constantly change directories and / or open new images. It operates by recognizing all the figures in a folder and bringing up each figure automatically allowing the user to click and enter relevant information about a figure as they go. All this information is then stored in a data frame or list at the end of the process, saving quite a bit of time. Users can stop mid-way through a folder by simply exiting after the last plot they have digitised. The data from completed figures will automatically be written to a special `caldat/` folder that is created when first executing the metaDigitise function. Digitised data along with metadata for figures are stored in caldat/ for later use and editing (see below)

In order to correctly extract data from figures metaDigitise always requires the user to calibrate the axes in the figure. To do this, the user is required to click on two known points on the axis in question, and then enter the value of those points in the figure (Figure \ref{fig:calibrate}). Using this information, metaDigitise then calculates the value of any clicked points in terms of the figure axes. In the case of mean and error plots and box plots, it calibrates only the y axis (assuming the x axis is redundant). For scatter plots and histograms both axes are calibrated.

@@

A particularly powerful and flexible aspect of metaDigitise is its ability to recognize figures that are not already completed and only bring up those figures that have not been digitised in subsequent calls of the function. This means that new figures can be added at any time as the project develops and previously digitised figures are simply ignore during processing, but their data re-integrated within the final output after new files are completed.

From all plot types, metaDigitise summarises the data from a figure or set of figures as a mean, standard deviation and sample size, for each identified group within the plot (should multiple groups exist). These are the summary statistics needed to create many of the relevant effect sizes and sampling error for a meta-analysis. In the case of scatter plots, metaDigitise also returns the correlation coefficient between the points within each identified group.

\subsection{Diverse Plot Types}
metaDigitise recognises four main types of plot; Mean and error plots, box plots, scatter plots and histograms, shown in Figure \ref{fig:plot_type}. Each of these can be processed together and integrated into a single output. Alternatively, users can keep like figures together and process them separately.

%%% Equations for mean and sd
%% pooled stats: 
% \begin{align}
% \mu _{X}&={\frac {\sum _{i}{N_{X_{i}}\mu _{X_{i}}}}{\sum _{i}{N_{X_{i}}}}}
% \\
% \sigma _{X}&={\sqrt {{\frac {1}{\sum _{i}{N_{X_{i}}-1}}}\left(\sum _{i}{\left[(N_{X_{i}}-1)\sigma _{X_{i}}^{2}+N_{X_{i}}\mu _{X_{i}}^{2}\right]}-\left[\sum _{i}{N_{X_{i}}}\right]\mu _{X}^{2}\right)}}
% \end{align}

@@

\subsection{Re-importing, Editing and Plotting Previously Digitised data}

A particularly useful feature of `metaDigitise` is its ability to re-import, edit and re-plot previously digitised figures. Running metaDigitise brings up a series of prompts for the user using a main menu that provides access to a number of its features:

\begin{verbatim}
	Do you want to...
1: Process new images
2: Import existing data
3: Edit existing data
Selection: 
\end{verbatim}

The user simply enters in the numeric value that corresponds to what they would like to do. Choosing "Process new images" will permit users to digitise new figures, integrating previously completed digitisations along with newly digitised data together at the end of the session, or when the user decides to quit. If the user chooses "Import existing data", they have the option of either 1) importing all data or 2) importing a particular figure that has been digitised. If 2, then a list of files are provided to the user that they can select.

Editing existing data allows users to easily re-plot or edit information or digitisations that have previously be done for any plot. This is accomplished by guiding the user through a new set of options:

\begin{verbatim}
Choose how you want to edit files:
1: Cycle through images
2: Choose specific file to edit
3: Enter previously omitted sample sizes
Selection: 
\end{verbatim}

If the user is unsure about the name of the specific figure they need to edit or simply want to just check the digitisations of figures they can choose "Cycle through images", which will bring up each figure, one by one, overlaying the calibrations, group names (if they exist), sample sizes (if they were entered) and the selected points. Alternatively, choosing option 2, will bring up a list of the completed files in the folder and the specific file can be chosen, at which point it will be replotted. Either of these options will cycle through a number of questions asking the user what they would like to edit:

\begin{verbatim}
Edit rotation? If yes, then the whole extraction will be redone (y/n) n
Change plot type? If yes, then the whole extraction will be redone (y/n) n
Variable entered as: Sepal length
Rename Variables (y/n) n
Edit calibration? (y/n) n
Re-extract data (y/n) y
Change group identifier? (y/n) n
Add group, Delete group or Finish plot? a/d/f d
1: setosa
2: versicolor
3: viriginica
Selection: 2
Add group, Delete group or Finish plot? a/d/f a
\end{verbatim}

A whole host of information can be edited from the rotation, plot type, the variable name that was provided, the calibration and even the digitisation of groups. When editing the metaDigitise object is re-written to the caldat folder and the edits are immediately integrated into the existing object once complete.

\section{Additional Features}

\subsection{Figure Rotation and Adjustment}
Figure may have been extracted from old publications, for example from scanned images, and so are not perfectly orientated on the image. This will make the calibration of the points in the figure from the image problematic. metaDigitise allows users to rotate the image. By clicking two points on the x-axis, metaDigitse calculates the angle needed to rotate the image so the x-axis is horizontal, and rotates it. (Figure \ref{fig:rotate}A,B)

\subsection{Obtaining Raw Data}

While metaDigitise provides users with the summary statistics by default, for all plot types, in many cases the user may actually be interested in obtaining the raw digitised data from scatter plots. This is very easy to do my changing the default `summary` argument from TRUE to FALSE in the metaDigitise function. Instead of providing the user with summary statistics it will return a list containing four slots for each of the figure types (mean error, box plot, histogram and scatter plots). An example of a data object returned from digitising figures is as follows:

\begin{verbatim}
str(data)
List of 3
 $ mean_error :List of 1
  ..$ 002_Doe_2013_Fig1.png:'data.frame': 3 obs. of  5 variables:
  .. ..$ id      : Factor w/ 3 levels "setosa","versicolor",..: 1 2 3
  .. ..$ mean    : num [1:3] 5 5.93 6.59
  .. ..$ error   : num [1:3] 0.111 0.148 0.178
  .. ..$ n       : num [1:3] 50 50 50
  .. ..$ variable: chr [1:3] "Sepal length" "Sepal length" "Sepal length"
 $ hist       :List of 1
  ..$ 003_Doe_2013_Fig3.png:'data.frame': 8 obs. of  3 variables:
  .. ..$ midpoints: num [1:8] 4.3 4.5 4.7 4.9 5.1 ...
  .. ..$ frequency: num [1:8] 4 5 7 12 11 6 2 3
  .. ..$ variable : chr [1:8] "Sepal length" "Sepal length" "Sepal length" "Sepal length" ...
 $ scatterplot:List of 1
  ..$ 001_Anderson_1935_Fig1.png:'data.frame':  83 obs. of  8 variables:
  .. ..$ id        : Factor w/ 2 levels "setosa","versicolor": 1 1 1 1 1 1 1 1 1 1 ...
  .. ..$ x         : num [1:83] 2.3 2.9 3 3 3 ...
  .. ..$ y         : num [1:83] 4.5 4.4 4.41 4.3 4.8 ...
  .. ..$ group     : num [1:83] 1 1 1 1 1 1 1 1 1 1 ...
  .. ..$ col       : Factor w/ 2 levels "red","green": 1 1 1 1 1 1 1 1 1 1 ...
  .. ..$ pch       : num [1:83] 19 19 19 19 19 19 19 19 19 19 ...
  .. ..$ y_variable: chr [1:83] "Sepal length (mm)" "Sepal length (mm)" "Sepal length (mm)" "Sepal length (mm)" ...
  .. ..$ x_variable: chr [1:83] "Sepal width (mm)" "Sepal width (mm)" "Sepal width (mm)" "Sepal width (mm)" ...
\end{verbatim}

Here, the user can easily access the list of raw scatter plot data by simply extracting the scatter plot slot:

\begin{verbatim}
scatterplot <- data$scatterplot
\end{verbatim}

%check extraction
%Including addition of N later

\section{Example}
%% scatterplot from airquality data
We'll now turn to an example of how metaDigitise works using figures generated using the well known iris data set \ref{fig:all_extract}. Users can install metaDigitise from GitHub as follows:

Assume that we have three figures, from two hypothetical studies, that we would like to extract summary statistics from. Each hypothetical study quantifies sepal length and width on a number of iris species.
